A weakly structured stem for human origins in Africa

### 2022-07-22 paper.tex
Update paper.tex

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological record of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features were intermittantly present across the continent
300-100ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic hominin admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, continuous structure, or hybrid models including both weak structure and archaic
hominin admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic hominin admixture model from the weakly-structured-stem model.

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of
archaic hominin admixture having also occurred in Africa. For this reason, previous
work has focused on archaic hominin admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that
weakly-structured-stem models better capture these patterns.
They also help explain an
ecological riddle posed by the archaic hominin admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
ancestral human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly-structured-stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

>>>>

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological records of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features were intermittantly present across the continent
300-100ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic hominin admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, continuous structure, or hybrid models including both weak structure and archaic
hominin admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic hominin admixture model from the weakly-structured-stem model.

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of
archaic hominin admixture having also occurred in Africa. For this reason, previous
work has focused on archaic hominin admixture to explain patterns of polymorphism
inconsistent with a single-origin model. Here, we have shown that
weakly-structured-stem models better capture these patterns.
They also help explain an
ecological riddle posed by the archaic hominin admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
ancestral human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly-structured-stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.
